Concurrent Programming with CXL

### 2020-06-29 book/paper.tex
Made formatting pass over appendices

Tuples, lists, strings, and bags are all special cases of dictionaries.
Both tuples and lists map indexes (starting at 0) to CXL values.
Their format is either \texttt{(e, e, ..., e,)} or
\texttt{[e, e, ..., e,]}.
If the tuple or list has two or more elements, then the final comma
is optional.
A string must be enclosed by double quotes and is represented as
a tuple of its characters.  Characters are one-character atoms.
A bag or multiset is a dictionary that maps a value to how many
times it occurs in the bag.

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  \texttt{None} is the empty list of atoms.
Addresses are lexicographically ordered accordingly;
\item contexts (Appendix~\ref{app:context}) are ordered by hash.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{e == e}, \texttt{e != e} & two CXL values of the same type \\
\hline
\texttt{e < e, e <= e} & any two CXL values \\
\hline
\texttt{e > e, e >= e} & any two CXL values \\
\hline
\texttt{e and e and ...} & two or more booleans \\
\hline
\texttt{e or e or ...} & two or more booleans \\
\hline
\texttt{not e} & a boolean \\
\hline
\texttt{-e} & an integer \\
\hline
\texttt{e + e + ...} & two or more integers, sets, or lists \\
\hline
\texttt{e - e} & two integers or sets \\
\hline
\texttt{e * e * ...} & two or more integers or sets \\
\hline
\texttt{e / e} & two integers (integer division) \\
\hline
\texttt{e \% e} & two integers (division remainder) \\
\hline
\texttt{e~..~e} & two integers \\
\hline
\texttt{e [not] in e} & first is any CXL value, second is set \\
\hline
\texttt{e if e else e} & middle \texttt{e} must be a boolean \\
\hline
\texttt{\^{}e} & an address \\
\hline
\texttt{choose e} & a set \\
\hline
\texttt{min e, max e} & a set \\
\hline
\texttt{cardinality e} & a set \\
\hline
\texttt{keys e} & a dictionary \\
\hline
\texttt{len e} & a dictionary (includes tuple or list) \\
\hline
\texttt{bagsize e} & a bag \\
\hline
\texttt{atLabel e} & atom (corresponding to a label) \\
\hline
\texttt{nametag e} & \texttt{e} must be \texttt{()} \\
\hline
\texttt{processes e} & \texttt{e} must be \texttt{()} \\
\hline
\texttt{\&lv} & \texttt{lv} is an lvalue \\
\hline
\texttt{stop lv} & \texttt{lv} is an lvalue \\
\hline
\end{tabular}
\vspace{1em}

\texttt{+} computes the union when applied to sets and
the concatenation when applied to lists or tuples.
\texttt{-} computes set difference when applied to two sets.
\texttt{*} computes the intersection when applied to sets.
\texttt{x..y} computes the set consisting of the integers \texttt{x}
through \texttt{y} inclusive.  It returns the empty set if \texttt{x > y}.

An \emph{lvalue}
\index{lvalue}%
(short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a process variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\texttt{var}, \texttt{var[e]}, and \texttt{\^{}p}.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Set comprehension & \texttt{\{ f(e) for e in s \}} \\
\hline
List comprehension & \texttt{[ f(e) for e in s ]} \\
\hline
Dict comprehension & \texttt{dict\{ f(e) for e in s \}} \\
\hline
\end{tabular}
\vspace{1em}

In each case, \texttt{s} must be a set.
For list comprehensions, the list will be sorted by \texttt{e}.
For dictionary comprehensions, each \texttt{e} will be mapped to \texttt{f(e)}.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression\\
\hline
\texttt{lv [op]= e} & op is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
\texttt{and}, and \texttt{or}\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del lv} & delete\\
\hline
\texttt{assert b [, x]} & \texttt{b} is a boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
\hline
\texttt{const v = e} & \texttt{v} is an identifier, \texttt{e} is a constant expression\\
\hline
\texttt{def m(v, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
\hline
\texttt{let v = e, ...: S} & \texttt{e} is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if b: S} & \texttt{b} is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{while b: S} & \texttt{b} is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{for v in e: S} & \texttt{e} is a set evaluated in order,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn p e [, t]} & \texttt{p} is a method, \texttt{e} is an expression, \texttt{t} is an optional tag (an expression) \\
\hline
\texttt{go c e} & \texttt{c} is a context, \texttt{e} is an expression \\
\hline
\texttt{import m} & \texttt{m} identifies a module \\
\hline
\end{tabular}
\vspace{1em}

CXL does not support Python-style iterators.
To iterate through a dictionary, list, or tuple, use:
\texttt{for v in keys(e):~S}.

@@

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{recAlloc()} & allocate a record \\
\hline
\texttt{recFree(r)} & free record \texttt{r} \\
\hline
\end{tabular}
\vspace{1em}

@@

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{bagEmpty()} & returns an empty bag\\
\hline
\texttt{bagFromSet(s)} & create a bag from set \texttt{s}\\
\hline
\texttt{bagCount(b, e)} & count how many times \texttt{e} occurs in bag \texttt{b}\\
\hline
\texttt{bagChoose(b)} & like \texttt{choose(s)}, but applied to a bag\\
\hline
\texttt{bagAdd(pb, e)} & add one copy of \texttt{e} to bag \texttt{\^{}pb}\\
\hline
\texttt{bagRemove(pb, e)} & remove one copy of \texttt{e} from bag \texttt{\^{}pb}\\
\hline
\end{tabular}
\vspace{1em}

\noindent
\index{list module}%
The \texttt{list} module has various useful methods that operate on lists
or tuples:

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{subseq(t, b, f)} & return a \emph{slice} of list \texttt{t} starting
at index \texttt{b} and ending just before index \texttt{f}\\
\hline
\texttt{append(t, e)} & append \texttt{e} to list \texttt{t}\\
\hline
\texttt{head(t)} & return the first element of list \texttt{t}\\
\hline
\texttt{tail(t)} & return all but the first element of list \texttt{t}\\
\hline
\texttt{listQsort(t)} & sort list \texttt{t} using quicksort\\
\hline
\texttt{list2bag(t)} & convert list \texttt{t} into a bag \\
\hline
\texttt{list2set(t)} & convert list \texttt{t} into a set \\
\hline
\texttt{listMin(t)} & return the minimum of all elements in \texttt{t}\\
\hline
\texttt{listMax(t)} & return the maximum of all elements in \texttt{t}\\
\hline
\texttt{listSum(t)} & return the sum of all elements in \texttt{t}\\
\hline
\end{tabular}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Address $n$ & compute address from $n$ components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert & pop $b$ and check that it is \texttt{True} \\
\hline
AtomicInc & increment the atomic counter of this context \\
\hline
AtomicDec & decrement the atomic counter of this context \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete process variable $v$ \\
\hline
Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame $m$($a$) & start method $m$ with arguments $a$,
initializing variables.  Last instruction is at $r$ \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a process variable onto the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
PushAddress $v$ & push the address of variable $v$ \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Set & pop $n$ and $n$ elements and push a set of the elements \\
\hline
Spawn & pop tag, argument, and method and spawn a new context \\
\hline
Split & pop set and push minimum and remainder \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
\hline
Swap & swap the two values on top of the stack \\
\hline
\end{tabular}
\vspace{1em}

>>>>

Tuples, lists, strings, and bags are all special cases of dictionaries.
Both tuples and lists map indexes (starting at 0) to CXL values.
Their format is either \texttt{($e$, $e$, ..., $e$,)} or
\texttt{[$e$, $e$, ..., $e$,]}.
If the tuple or list has two or more elements, then the final comma
is optional.
A string must be enclosed by double quotes and is represented as
a tuple of its characters.  Characters are one-character atoms.
A bag or multiset is a dictionary that maps a value to how many
times it occurs in the bag.

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  \texttt{None} is the empty list of atoms.
Addresses are lexicographically ordered accordingly;
\item contexts (Appendix~\ref{app:context}) are ordered first by name tag, then by program counter, then by hash.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two CXL values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$} & any two CXL values \\
\hline
\texttt{$e$ > $e$, $e$ >= $e$} & any two CXL values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, sets, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ * $e$ * ...} & two or more integers or sets \\
\hline
\texttt{$e$ / $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$} & two integers (division remainder) \\
\hline
\texttt{$e$~..~$e$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any CXL value, second is set \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{\^{}$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$} & a set \\
\hline
\texttt{cardinality $e$} & a set \\
\hline
\texttt{keys $e$, len $e$} & a dictionary (includes tuple or list) \\
\hline
\texttt{bagsize $e$} & a bag \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{processes $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{\&}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\end{tabular}
\vspace{1em}

\texttt{+} computes the union when applied to sets and
the concatenation when applied to lists or tuples.
\texttt{-} computes set difference when applied to two sets.
\texttt{*} computes the intersection when applied to sets.
\texttt{$x$..$y$} computes the set consisting of the integers $x$
through $y$ inclusive.  It returns the empty set if $x > y$.

An \emph{lvalue}
\index{lvalue}%
(short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a process variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\textit{var}, \textit{var}[$e$], and \texttt{\^{}}\textit{p}.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Set comprehension & \texttt{\{ f($e$) for $e$ in $s$ \}} \\
\hline
List comprehension & \texttt{[ f($e$) for $e$ in $s$ ]} \\
\hline
Dict comprehension & \texttt{dict\{ f($e$) for $e$ in $s$ \}} \\
\hline
\end{tabular}
\vspace{1em}

In each case, $s$ must be a set.
For list comprehensions, the list will be sorted by $e$.
For dictionary comprehensions, each $e$ will be mapped to \texttt{f}($e$).

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\textit{lv} = $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
\texttt{and}, and \texttt{or}\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{assert $b$ [, $x$]} & $b$ is a Boolean expression.  Optionally report value of expression $e$\\
\hline
\texttt{const v = $e$} & \texttt{v} is an identifier, $e$ is a constant expression\\
\hline
\texttt{def m($v$, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
\hline
\texttt{let $v$ = $e$, ...: S} & $v$ is an identifier, $e$ is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
\hline
\texttt{while $b$: S} & $b$ is a Boolean expression, \texttt{S} a list of statements\\
\hline
\texttt{for $v$ in $e$: S} & $v$ is an identifier, $e$ is a set evaluated in order,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is a tag (an expression) \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{import m} & \texttt{m} identifies a module \\
\hline
\end{tabular}
\vspace{1em}

CXL does not support Python-style iterators.
To iterate through a dictionary, list, or tuple, use:
\texttt{for $v$ in keys($e$):~S}.

@@

\section{The \texttt{alloc} module}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{recAlloc()} & allocate a record \\
\hline
\texttt{recFree($r$)} & free record $r$ \\
\hline
\end{tabular}
\vspace{1em}

@@

\section{The \texttt{bag} module}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{bagEmpty()} & returns an empty bag\\
\hline
\texttt{bagFromSet($s$)} & create a bag from set $s$\\
\hline
\texttt{bagCount($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
\hline
\texttt{bagChoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
\hline
\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{\^{}}\textit{pb}\\
\hline
\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{\^{}}\textit{pb}\\
\hline
\end{tabular}

\section{The \texttt{list} module}

\index{list module}%
The \texttt{list} module has various useful methods that operate on lists
or tuples:

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{subseq($t$, $b$, $b$)} & return a \emph{slice} of list $t$ starting
at index $b$ and ending just before index $f$\\
\hline
\texttt{append($t$, $e$)} & append $e$ to list $t$\\
\hline
\texttt{head($t$)} & return the first element of list $t$\\
\hline
\texttt{tail($t$)} & return all but the first element of list $t$\\
\hline
\texttt{listQsort($t$)} & sort list $t$ using quicksort\\
\hline
\texttt{list2bag($t$)} & convert list $t$ into a bag \\
\hline
\texttt{list2set($t$)} & convert list $t$ into a set \\
\hline
\texttt{listMin($t$)} & return the minimum of all elements in $t$\\
\hline
\texttt{listMax($t$)} & return the maximum of all elements in $t$\\
\hline
\texttt{listSum($t$)} & return the sum of all elements in $t$\\
\hline
\end{tabular}

\section{The \texttt{synch} module}

\index{synch module}%
The \texttt{synch} module provides the following methods:

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
tas(\textit{lk}) & test-and-set on \texttt{\^{}}\textit{lk} \\
\hline
Lock() & return a lock object \\
\hline
lock(\textit{lk}) & acquire \texttt{\^{}}\textit{lk} \\
\hline
unlock(\textit{lk}) & release \texttt{\^{}}\textit{lk} \\
\hline
Condition(\textit{lk}) & return a condition variable on \texttt{\^{}}\textit{lk} \\
\hline
wait($c$) & wait on condition variable \texttt{\^{}}$c$ \\
\hline
notify($c$) & notify a process waiting on condition variable \texttt{\^{}}$c$ \\
\hline
notifyAll($c$) & notify all processes waiting on condition variable \texttt{\^{}}$c$ \\
\hline
Semaphore(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
\hline
P(\textit{sema}) & procure \texttt{\^{}}\textit{sema}  \\
\hline
V(\textit{sema}) & vacate \texttt{\^{}}\textit{sema}  \\
\hline
Queue() & return a synchronized queue object \\
\hline
dequeue($q$) & return next element of $q$, blocking if empty \\
\hline
enqueue($q$, \textit{item}) & add \textit{item} to $a$ \\
\hline
\end{tabular}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Address $n$ & compute address from $n$ components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert & pop $b$ and check that it is \texttt{True} \\
\hline
AtomicInc & increment the atomic counter of this context \\
\hline
AtomicDec & decrement the atomic counter of this context \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete process variable $v$ \\
\hline
Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m}($a$) & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a process variable onto the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
PushAddress $v$ & push the address of variable $v$ \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Set & pop $n$ and $n$ elements and push a set of the elements \\
\hline
Spawn & pop tag, argument, and method and spawn a new context \\
\hline
Split & pop set and push minimum and remainder \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a process variable \\
\hline
Swap & swap the two values on top of the stack \\
\hline
\end{tabular}
\vspace{1em}
